Amphistomy increases leaf photosynthesis more in coastal than montane plants of Hawaiian ʻilima (*Sida fallax*)

### 2023-11-03 ms/ms.qmd
revised ms

\textbf{Conclusions}  This is the first direct experimental evidence that amphistomy *per se* increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase the supply of CO$_2$ to chloroplasts. The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in ʻsunʻ leaves, but the mechanistic basis of this observation is an area for future research.

@@

\autoref{fig:ags-curve} illustrates our method for collecting $A \textendash g_\text{sw}$ curves. We collected two curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve. To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements. In the field, we acclimated the focal leaf to saturating light and high relative humidity ($\mathrm{RH} = 70\%$), as described above, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process of acclimating the leaf to 70\% $\mathrm{RH}$ and inducing stomatal closure with low $\mathrm{RH}$ with the other treatment (amphi or pseudohypo). We did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown).

@@

In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap between the curves. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. The vertical lines in \autoref{fig:licor} show the $g_\text{sw}$ for each leaf. $\mathrm{AA}$ for each leaf was calculated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ using the log-response ratio shown above.

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants more than that of montane plants. $\mathrm{AA}$ was significantly greater than 0 (95% confidence intervals did not overlap 0) in 5 of 6 coastal leaves, but only 1 of 6 montane leaves (\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual curves). Overall, the average $\mathrm{AA}$ among coastal and montane leaves is $`r AA_coastal`$ and $`r AA_montane`$, respectively; the difference in average $\mathrm{AA}$ between habitat types is $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = `r AA_diff`$. Posterior predictions closely match observed values of $A$ (\autoref{fig:pp-licor}), indicating an adequate model fit from which we can interpolate between measurements reliably. It also suggests that slight extrapolation beyond the data should be reliable, but this is less certain. When we remove two leaves where extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.  Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}). Water-use efficiency ($A / g_\mathrm{sw}$) was significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['wue']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['wue']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['wue']][['p.value']], 2)`$).

@@

Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why. The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* affects improves photosynthesis in a given leaf. By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf. This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant. Taking advantage of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs. $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$). Coastal and montane ʻilima leaves are likely good representatives of classic sun and shade leaf syndromes because 1) they vary in traits like reflective pubescence [@ehleringer_pubescence_1978] and leaf thickness [@terashima_why_2001] that typically characterize sun-shade adaptation; and 2) since ʻilima shrubs are typically $<1$m tall, they are shaded by trees in montane, but not coastal habitats (\autoref{fig:study-system}d). While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in this environment.

If $\mathrm{AA}$ is typically greater in sun leaves than shade leaves, it could partially explain the distribution of amphi- and hypostoamtous leaves, but the precise mechanism(s) require further study. One hypothesis is that the internal airspace conductance, $g_\mathrm{ias}$, from stomata to mesophyll cell walls is lower in thicker sun leaves [@parkhurst_adaptive_1978]. All else being equal, a leaf with lower $g_\mathrm{ias}$ will benefit more from amphistomy. Our results partially support this hypothesis. Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}a), opposite our prediction. Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats. Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves. The negative relationship, (albeit nonsignificant in that 95\% confidence intervals encompassed 0), between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].

@@

Among land plants grown in a common garden, amphistomatous leaves have on average nearly $2\times$ higher area-based photosynthetic rates [@xiong_one_2020], naively implying an $\mathrm{AA} \approx \text{log}~2 = 0.69$. This is much higher than our estimate of `r round(habitat_aa$coastal, 2)` among coastal ʻilima leaves. The most likely explanation is that amphistomy is not the only cause of high photosynthetic rate. Indeed, species adapted to open, high light habitats with amphistomatous leaves also have higher concentrations of Rubisco, overall stomatal conductance, and photosynthetic capacity [@smith_leaf_1997; @xiong_one_2020]. For a leaf with high photosynthetic capacity that is well illuminated and hydrated, the major limitation becomes CO$_2$. Under these conditions, amphistomy may substantially increase photosynthesis, as we observe in coastal ʻilima leaves. Selection on increased photosynthesis under similar conditions may explain why crops leaves tend to increase stomatal density ratio during domestication [@milla_shifts_2013].

Three limitations of this study are the small sample size, experimental design that precludes distinguishing genetic from environmental differences in leaf traits, and potentially confounding effects of other environmental differences besides light environment. Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes. Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008]. Each of these factors and others potentially modulate $\mathrm{AA}$. Quantifying the contribution of all these factors requires larger samples and additional measurements that are beyond the scope of this study, but exciting avenues for future research on leaf structure-function relations. Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*. While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be insightful to know about genetic and environmental contributions to trait variation. A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature. However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature and precipitaiton. Experimental studies in controlled environments will be necessary to isolate the effect light quantity and quality on $\mathrm{AA}$.

>>>>

\textbf{Conclusions}  This is the first direct experimental evidence that amphistomy *per se* increases photosynthesis, consistent with the hypothesis that parallel pathways through upper and lower mesophyll increase the supply of CO$_2$ to chloroplasts. The prevalence of amphistomatous leaves in open, sunny habitats can partially be explained the increased benefit of amphistomy in `sun' leaves, but the mechanistic basis of this observation is an area for future research.

@@

\autoref{fig:ags-curve} illustrates our method for collecting $A \textendash g_\text{sw}$ curves. We collected two curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve. To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements, though we did not detect an effect of treatment order on $\mathrm{AA}$ (results not shown). In the field, we acclimated the focal leaf to saturating light and high relative humidity ($\mathrm{RH} = 70\%$), as described above, until $A$ and $g_\text{sw}$ reach their maximum. We used these data as our estimates of maximum $A$ and $g_\text{sw}$. After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation. Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity. All other environmental conditions in the leaf chamber remained the same. We logged data until $g_\text{sw}$ reached its nadir. We then repeated the process of acclimating the leaf to 70\% $\mathrm{RH}$ and inducing stomatal closure with low $\mathrm{RH}$ with the other treatment (amphi or pseudohypo).

@@

In 10 of 12 leaves, the minimum $g_\text{sw}$ of the amphi curve was smaller than the maximum $g_\text{sw}$ of the pseudohypo curve (i.e. the curves overlapped for a range of $g_\text{sw}$ values). In those cases, we estimated $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ at the $g_\mathrm{sw}$ value in the middle of the range of overlap between the curves. In 2 of 12 leaves, the $A \textendash g_\text{sw}$ curves did not quite overlap because the minimum $g_\text{sw}$ of the amphi curve was slightly greater than the maximum $g_\text{sw}$ of the pseudohypo curve. In those cases, we estimated $\mathrm{AA}$ by extrapolating slightly, `r gsw_extrap`, beyond the measured curves to the $g_\mathrm{sw}$ value in between the curves. The vertical lines in \autoref{fig:licor} show the $g_\text{sw}$ for each leaf. We estimated $\mathrm{AA}$ from $\hat{A}_\text{amphi}$ and $\hat{A}_\text{hypo}$ for each leaf using the log-response ratio shown above.

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants more than that of montane plants. $\mathrm{AA}$ was significantly greater than 0 (95% confidence intervals did not overlap 0) in 5 of 6 coastal leaves, but only 1 of 6 montane leaves (\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual curves). Overall, the average $\mathrm{AA}$ among coastal and montane leaves is $`r AA_coastal`$ and $`r AA_montane`$, respectively; the difference in average $\mathrm{AA}$ between habitat types is $\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = `r AA_diff`$. Posterior predictions closely match observed values of $A$ (\autoref{fig:pp-licor}), indicating an adequate model fit from which we can interpolate between measurements reliably. It also suggests that slight extrapolation beyond the data should be reliable, but this is less certain. When we remove two leaves where we extrapolated slightly beyond fitted $A \textendash g_\mathrm{sw}$ curves, we estimate that $\mathrm{AA}_\text{coastal}$ is still positive, $`r AA_coastal1`$, but the difference between coastal and montane leaves is smaller,  $`r AA_diff1`$, and confidence intervals slightly overlap 0.  Maximum photosynthetic rate was slightly, but not significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['A']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['A']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['A']][['p.value']], 2)`$); total stomatal conductance was similar (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['gsw']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['gsw']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['gsw']][['p.value']], 2)`$) in coastal and montane leaves (\autoref{fig:habitat-Ags}). Water-use efficiency ($A / g_\mathrm{sw}$) was significantly higher in coastal leaves (Welch Two Sample *t*-test, $t_{`r round(fit_habitat_Ags[['wue']][['parameter']], 2)`} = `r round(fit_habitat_Ags[['wue']][['statistic']], 2)`$, $P = `r round(fit_habitat_Ags[['wue']][['p.value']], 2)`$).

@@

Amphistomy is a seemingly simple way that leaves can increase carbon gain without significant additional water loss, yet it is rare in nature and we do not know why. The strong association between amphiostomy and sunny, open habitats suggests that amphistomy may benefit sun leaves more than shade leaves, but progress has been limited by the lack of evidence that amphiostomy *per se* improves photosynthesis in a given leaf. By experimentally blocking gas exchange through the upper surface in a controlled environment, we directly compared an amphistomatous leaf to an otherwise identical pseudohypostomatous leaf. This allows us to quantity the amphistomy advantage ($\mathrm{AA}$) holding all else constant. Taking advantage of the steep climatic gradients in the Hawaiian archipelago, we applied this new method to show for the first time that sun leaves benefit  `r prettyNum(habitat_aa$times_increase, digits = 3L)` times more from amphistomy than shade leaves on ʻilima (*Sida fallax*) plants ($\mathrm{AA}_\text{coastal} = `r prettyNum(habitat_aa[['coastal']], digits = 2L)`$ vs. $\mathrm{AA}_\text{montane} = `r prettyNum(habitat_aa[['montane']], digits = 2L)`$). Coastal and montane ʻilima leaves are likely good representatives of classic sun and shade leaf syndromes because 1) they vary in traits like reflective pubescence [@ehleringer_pubescence_1978] and leaf thickness [@terashima_why_2001] that typically characterize sun-shade adaptation; and 2) since ʻilima shrubs are typically $<1$m tall, they are shaded by trees in montane, but not coastal habitats (\autoref{fig:study-system}d). While this result has not yet been validated in other species, our results indicate that part of the reason amphistomatous leaves are found most commonly in high light habitats is that the adaptive benefit is greater in such environments.

If $\mathrm{AA}$ is typically greater in sun leaves than shade leaves, it could partially explain the distribution of amphi- and hypostoamtous leaves, but the precise mechanism(s) require further study. One hypothesis is that the internal airspace conductance, $g_\mathrm{ias}$, from stomata to mesophyll cell walls is lower in thicker sun leaves [@parkhurst_adaptive_1978]. All else being equal, a leaf with lower $g_\mathrm{ias}$ will benefit more from amphistomy. Our results partially support this hypothesis. Coastal ʻilima leaves with high $\mathrm{AA}$ (\autoref{fig:habitat-aa}) are thicker than montane leaves, but the relationship between $\mathrm{AA}$ and leaf thickness within habitats is actually slightly negative (\autoref{fig:traits-aa}a), opposite our prediction. Since coastal and montane leaves differ in many respects besides thickness, we do not have enough data to conclude that leaf thickness explains the variation in $\mathrm{AA}$ between habitats. Alternatively, other biochemical or anatomical differences between coastal and montane leaves may explain why $\mathrm{AA}$ is greater in coastal leaves. The negative relationship, albeit nonsignificant in that 95\% confidence intervals encompassed 0, between leaf thickness and $\mathrm{AA}$ could be explained if thicker leaves compensated by having a more porous mesophyll and/or less tortuous airspaces [@theroux-rancourt_maximum_2021].

@@

Among land plants grown in a common garden, amphistomatous leaves have on average nearly $2\times$ higher area-based photosynthetic rates [@xiong_one_2020], naively implying an $\mathrm{AA} \approx \text{log}~2 = 0.69$. This is much higher than our estimate of `r round(habitat_aa$coastal, 2)` among coastal ʻilima leaves. The most likely explanation is that amphistomy is not the only cause of high photosynthetic rate. Indeed, species adapted to open, high light habitats with amphistomatous leaves also have higher concentrations of Rubisco, overall stomatal conductance, and photosynthetic capacity [@smith_leaf_1997; @xiong_one_2020]. For a leaf with high photosynthetic capacity that is well illuminated and hydrated, the major limitation becomes CO$_2$. Under these conditions, amphistomy may substantially increase photosynthesis, as we observe in coastal ʻilima leaves. Selection on increased photosynthesis under similar conditions may explain why crop leaves tend to increase stomatal density ratio during domestication [@milla_shifts_2013].

Three limitations of this study are the small sample size, experimental design that precludes distinguishing genetic from environmental differences in leaf traits, and potentially confounding effects of other environmental differences besides light environment. Understanding the mechanistic basis of higher $\mathrm{AA}$ in sun leaves would require much larger sample sizes. Sun leaves tend to be thicker, more densely packed with mesophyll cells, and have greater photosynthetic capacity and higher stomatal conductance, among other traits [@lambers_plant_2008]. Each of these factors and others potentially modulate $\mathrm{AA}$. Quantifying the contribution of all these factors requires larger samples and additional measurements that are beyond the scope of this study, but exciting avenues for future research on leaf structure-function relations. Although many morphological traits that distinguish coastal and montane ʻilima populations persist in a common environment [@yorkston_interfertility_2006], we cannot distinguish between genetic effects and plastic responses to habitat as causes of difference in $\mathrm{AA}$ because we measured naturally occurring plants *in situ*. While disentangling genetic and plastic contributions is not necessarily important for understanding the distribution of amphistomatous leaves, it would be insightful to know about genetic and environmental contributions to trait variation. A reciprocal transplant would be able to determine the genetic and environmental contributions, as well their interaction, to trait variance in nature. However, reciprocal transplants cannot control for other differences between coastal and montane habitats besides vegetation height, such as temperature and precipitaiton. Experimental studies in controlled environments will be necessary to isolate the effects of light quantity and quality on $\mathrm{AA}$.
